Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-06-07 joint_inf.tex
I think I fixed it?

This split formulation defines mappings from patterns to splits as
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
Given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define random variables
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV)
that take corresponding realizations $\siteSplit_j$ and $\ancestralSplit_k$ for some $j$ and $k$.
We write the $i$th factor of \eqref{eq:full_likelihood} as
%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
for some $j$ and $k$.
Taking the maximum across internal states yields
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

>>>>

This split formulation defines mappings from patterns to splits as
 $$
 \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
 $$
Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$.

%EM Wait, is this first statement actually true? It seems impossible because multiple y_i's collaps to a single psi(y_i). At least it needs a factor of 2, but it seems more complex than that.
The $i$th factor of \eqref{eq:full_likelihood} has the property
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we need to define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
that takes both the ancestral state column and site column as input.
Given $\alignmentColumnRV$ as well as an ancestral state--valued random variable $\ancestralStateColumnRV$ define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
$$
Then
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                                                                                                           &= P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ and $k$.

Taking the maximum across internal states yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.
